---
title: "Oligodendrocytes genes clustering"
author: "Elena Brivio"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
    code_folding: hide

---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, fig.height = 12, fig.width = 12)
```
# Set-up
```{r, message=FALSE, warning=FALSE}
rm(list = ls())
library(tidyverse)
library(Seurat)
library(ggplot2)
library(tibble)
library(viridis)
library(ggrepel)
library(eulerr)
library(dendextend)
library(factoextra)

sessionInfo()
```


# Import Data
```{r}
fc <- read.csv("Genes/differential_genes/combined.csv")
fc <- fc %>% 
  dplyr::filter(cell_type == "Oligodendrocytes") %>% 
  separate(sex_exp, into = c("sex", "exp"), remove = F, sep = "_") 

DEG_list <- fc %>% filter(p_val_BH < 0.05) %>% pull(gene) %>% unique()

```


# Clustering genes on expression patterns
## Data Set-up
I select all gene, combining the 4 conditions and cluster based on the logFC regardless of significance
```{r}
set.seed(33)

fc_cl  <- fc %>% 
  filter(gene %in% DEG_list) %>% 
  select(gene, sex_exp, avg_logFC, p_val_BH) %>% 
  group_by(gene) %>% 
  complete(sex_exp, fill = list(avg_logFC = 0, p_val_BH = 1)) 
```
missing values are given a 0 as logFC and 1 as p val. 

Unclustered data
```{r}
ggplot(fc_cl, aes(y = reorder(gene, avg_logFC), x = sex_exp, fill = avg_logFC, color = p_val_BH < 0.05)) +
  geom_tile() +
  theme(axis.text.y = element_text(size = 2),
        plot.title = element_text(hjust = 0.5)) +
  scale_fill_gradient2(low = "#1C7987", mid = "white", high = "#9E2A2B", limits = c(min(fc_cl$avg_logFC), -min(fc_cl$avg_logFC))) +
  scale_color_manual(values = c("white", "black")) +
  ggtitle("Unclustered data")
```
>Only genes that are DEG in at least one condition are kept.

## Hierarchical clusterization of data - bottom-up
Transformation to matrix. Only values used for clustering (logFC) are retained
```{r}
hc <- fc_cl %>% 
  select(-p_val_BH) %>% 
  pivot_wider(id_cols = gene, names_from = sex_exp, values_from = avg_logFC) %>% 
  column_to_rownames("gene")
class(hc)
hc_df <- hc
hc <- as.matrix(hc)
```
Data Scaling
```{r}
hc <- scale(hc)
# head(hc)
```

## Distance calculation: Euclidean
```{r}
# distance matrix. euclidean distance.
dist_mat <- dist(hc, method = 'euclidean')
```


## Agnes clustering
Compare linkage methods and compute linkage coefficient. The higher agglomerative coefficient the better.
```{r}
m <- c( "average", "single", "complete", "ward")
names(m) <- c( "average", "single", "complete", "ward")
```

```{r}
purrr::map_dbl(m, function(x) {
  cluster::agnes(dist_mat, method = x)$ac
})
```
> Ward’s method gets us the highest agglomerative coefficient. I look at its dendogram.

```{r}
hclust_agnes <- cluster::agnes(dist_mat, method = "ward")
cluster::pltree(hclust_agnes, cex = 0.6, hang = -1, main = "Dendrogram of agnes")
```

### Dendrogram cutting
k = 16
Plot
```{r}
avg_dend_obj <- as.dendrogram(hclust_agnes)
plot.new()
avg_col_dend <- color_branches(avg_dend_obj, k = 16)
plot(avg_col_dend, #colored tree based on previous h
     leaflab = "none",
     main = "Gene Clustering",
     ylab = "Height")
#add the three cluster vectors
the_bars <- cutree(hclust_agnes, k = 16)
#this makes the bar
colored_bars(the_bars, avg_col_dend, sort_by_labels_order = T, y_shift=-0.2, rowLabels = "k = 16", cex.rowLabels=0.7, )
```

```{r}
cut_dendrogram <- cutree(hclust_agnes, k = 16)
```


## hclust Clustering
Linkage method chosen is Ward based on previous analysis
```{r}
hclust_hclust <- hclust(dist_mat, method = 'ward.D2')
plot(hclust_hclust)
```

### Dendrogram cutting
k = 16
Plot
```{r}
avg_dend_obj <- as.dendrogram(hclust_hclust)
plot.new()
avg_col_dend <- color_branches(avg_dend_obj, k = 16)
plot(avg_col_dend, #colored tree based on previous h
     leaflab = "none",
     main = "Gene Clustering",
     ylab = "Height")
#add the three cluster vectors
the_bars <- cutree(hclust_hclust, k = 16)
#this makes the bar
colored_bars(the_bars, avg_col_dend, sort_by_labels_order = T, y_shift=-0.2, rowLabels = "k = 16", cex.rowLabels=0.7, )

```

# Clusters plots

```{r}
# Create two dendrograms
dend1 <- as.dendrogram(hclust_agnes)
dend2 <- as.dendrogram(hclust_hclust)

tanglegram(dend1, dend2,
           highlight_distinct_edges = T, # Turn-off dashed lines
           common_subtrees_color_lines = FALSE, # Turn-off line colors
           common_subtrees_color_branches = F, # Color common branches 
          )
```

```{r}
sum(cut_dendrogram != the_bars)
```

> agnes and hclust create very similar but not identical dendograms, but the clustering is identical. 

> hclust is chosen.


```{r}
clValid::dunn(dist_mat, cut_dendrogram) 
```

> Dunn's index should be maximized, it represents clusters closed within and far outside (the ratio between the minimum inter-cluster distances to the maximum intra-cluster diameter).

```{r}
hc_df$cluster <- as.integer(cut_dendrogram)
```

```{r}
hc_df %>% group_by(cluster) %>% summarise(n = n())
```

# Plots

```{r}
hc_df1 <- hc_df %>%
  rownames_to_column("gene") %>% 
  pivot_longer(cols = c(female_ars, female_cms, male_ars, male_cms), names_to = "sex_exp", values_to = "logFC")
```


```{r, fig.height = 12, fig.width = 12}
hc_df1 %>% 
  ggplot(aes(x = sex_exp, y = logFC, group = gene, color = gene)) +
  geom_line(alpha = 0.5) +
  stat_summary(aes(group = cluster, fill = sex_exp),
               geom = "bar",
               fun = "median", color = "black", alpha = 0.7
               
  ) +
  facet_wrap(vars(cluster), scales = "free") +
  
  scale_fill_manual(values = col_degs) +
  theme_minimal() +
  theme(text = element_text(family = "Helvetica"),
        legend.position = "none", 
        axis.title.x = element_blank(), 
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 6),
        plot.title = element_text(hjust = 0.5)) 
```

```{r}
hc_df1 <- hc_df1 %>% 
  arrange(cluster) %>% 
  mutate(gene = factor(gene, levels = unique(gene))) 
```

# Heatmap

### Pheatmap
hclust dynamic cut
```{r}
my_gene_col <- cut_dendrogram
my_gene_col <- data.frame(my_gene_col) %>% rename("clusters"="my_gene_col") %>% 
  rownames_to_column("gene") %>% 
  mutate(clusters = factor(clusters)) %>% 
  column_to_rownames("gene")
colnames(hc) <- plyr::mapvalues(colnames(hc), from = colnames(hc), to = c("Baseline ARS\nFemale", "CMS ARS\nFemale", "Baseline ARS\nMale", "CMS ARS\n Male"))

set.seed(72)

paletteLength <- 100
colors_scale <- rev(hcl.colors(paletteLength, "RdBu"))
m <- max(abs(hc))

myBreaks <- c(seq(-m, 0, 
                  length.out = ceiling(paletteLength/2) + 1), 
              seq(m/paletteLength, m,
                  length.out = floor(paletteLength/2)))


pheatmap::pheatmap(mat = hc, 
                   scale = "none", 
                   clustering_method = "ward.D",
                   cluster_rows = hclust_hclust, 
                   cutree_rows = max(cut_dendrogram),
                   cluster_cols = F, 
                   annotation_row = my_gene_col,
                   color = colors_scale,
                   # annotation_colors = my_colour, 
                   fontsize_row = 2, 
                   legend_breaks = seq(-8, 8, 2),
                   legend_labels = seq(-8, 8, 2),
                   breaks = myBreaks, 
                   fontsize = 9, 
                   angle_col = 90, 
                   filename = paste0(output_images, "Pheatmap.png"),
                   width = 7, height = 8
                   )

```